---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# localgrid

<!-- badges: start -->
`r lifecycle::badge('experimental')`
<!-- badges: end -->

The goal of `localgrid` is to make it easy to define coordinate reference 
systems in the Oblique Mercator projection for new or existing arbitrary 
floating grids.

## Installation

You can install `localgrid` from Github with:

``` r
# install.packages("remotes")
remotes::install_github("crowcanyon/localgrid")
```

## Example

```{r example}
library(localgrid)

# Define a floating CRS for a new grid
define_new_grid(
    pt_0_geo = c(-108.51, 37.385),
    pt_0_grid = c(500, 500),
    name = "Great New Grid"
    )

# Or, define a floating CRS for a legacy grid
define_legacy_grid(
    pt_0_geo = c(-108.51, 37.385),
    pt_0_grid = c(500, 500),
    pt_1_geo = c(-108.511, 37.384),
    pt_1_grid = c(360, 380),
    name = "Screwy Legacy Grid"
    )
```

## Code of Conduct

Please note that the localgrid project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). 
By contributing to this project, you agree to abide by its terms.
