---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# localgrid

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Codecov test coverage](https://codecov.io/gh/crowcanyon/localgrid/branch/master/graph/badge.svg)](https://codecov.io/gh/crowcanyon/localgrid?branch=master)
[![R-CMD-check](https://github.com/crowcanyon/localgrid/workflows/R-CMD-check/badge.svg)](https://github.com/crowcanyon/localgrid/actions)
<!-- badges: end -->

The goal of `localgrid` is to make it easy to define coordinate reference 
systems in the Oblique Mercator projection for new or existing arbitrary 
floating grids.

## Installation

You can install `localgrid` from Github with:

``` r
# install.packages("remotes")
remotes::install_github("crowcanyon/localgrid")
```

## Example

```{r example}
library(localgrid)

# Define a floating CRS for a new grid
define_new_grid(
  pt_0_geo = c(-108.51, 37.385),
  pt_0_grid = c(500, 500),
  name = "Great New Grid"
)

# Or, define a floating CRS for a legacy grid
define_legacy_grid(
  pt_0_geo = c(-108.51, 37.385),
  pt_0_grid = c(500, 500),
  pt_1_geo = c(-108.511, 37.384),
  pt_1_grid = c(360, 380),
  name = "Screwy Legacy Grid"
)

# You can also use projected geographic points
define_legacy_grid(
  pt_0_geo = c(720446.8, 4140492.5),
  pt_0_grid = c(500, 500),
  pt_1_geo = c(720361.2, 4140379.2),
  pt_1_grid = c(360, 380),
  geo_crs = 6341,
  name = "Great New Grid"
)
```

## Code of Conduct

Please note that the localgrid project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). 
By contributing to this project, you agree to abide by its terms.
